\section{Discussion}

The purpose of this document is to describe an implementation of a
class structure for graphs (we define classes for both nodes and
edges) and consider some different problems that can arise.
This structure is based on GXL (reference: Holt and Winter).

There are many different ways to represent a graph and to deal with
the edges and nodes within that graph.
We would like to determine the operations that need to be performed on
a graph. These would essentially define the interface -- those
components that any one would need to implement to have access to
certain algorithms.

At the same time, we would like to define and implement a particular
class structure for graphs. This is somewhat orthogonal to the desire
to determine what the interface might be.

The class structure for nodes is:
\begin{description}
\item[id] An identifier (these must be unique), positive integers may
  be appropriate.
\item[type] A character vector describing the type of node.
\item[edgeOrder] The order that nodes should be drawn? Not sure about
  this particular one. It will be a vector of edge {\em id}'s.
\item[label] A label for the node. This is a character string.
\end{description}

The class structure for edges is:
\begin{description}
\item[id] An identifier (these must be unique); positive integers may
  be appropriate.
\item[type] A character vector describing the type of the edge.
\item[begin] The node {\em id} where the edge begins.
\item[end] The node {\em id} where the edge ends.
\end{description}

Both nodes and edges need to support attributes. These may be simply
name value pairs (this is the mechanism used by GXL).

It seems that relying on integers to index both nodes and edges
(separately) seems likely to help save on space and 

\section{Some Details}

Once both nodes and edges are instances of classes they will be quite
large. In order to reduce the storage requirements (especially for
large graphs) using integer indices may be beneficial.

The minimum amount of storage required is $|V|+|E|$. If we use an 
incidence matrix representation then the storage is $|V|^2$.
If we use a node and edge list representation then the storage
requirements are $|V|+2|E|$. When either $|V|$ or $|E|$ are large
these mechanisms will not be especially efficient.
In some cases it may be better to keep the actual node and edge data
stored in hash tables and keep other integer vectors available for
accessing the necessary components.

\section{Some Types of Graphs}

A brief digression on types of graphs:
\begin{description}
\item[digraph] A directed graph. All edges have direction (they go
  from one node to another but do not connect the nodes in the
  opposite direction).
\item[tree] A tree is an undirected acyclic graph (there are many
  other characterizations).
\end{description}

Some special data derived graph types:
\begin{description}
\item[cluster graph] A cluster graph is constructed when the nodes of
  the graph are clustered (partitioned). All nodes within a component
  of the partition are completely connected. There are no edges
  between components.
\end{description}


\section{Some operations on graphs}

Given two graphs $G_1$ and $G_2$ defined on the same set of nodes
we can ask about the intersection of these two graphs. This would be a
graph $G_3 = \{V, E^* \}$, where an edge is in $E^*$ if and only if it
is in both $E_1$ and $E_2$.

It will also be interesting to find the edges that are in $G_1$ say
that are not in $G_2$.

For a variety of the inferential problems it will be important to be
able to apply permutations to the graphs. It seems that often these
will require the graph structure itself to remain fixed and nodes
are simply relabled.

Other random constructs involve selecting edges between nodes in some
random fashion.

For trees (and especially for dendrograms) we might like to generate 
orders of the leaves that are consistent with the tree structure.
Possibly we could try to 

\subsection{Operations on Nodes}

We will want to find the inNodes and the outNodes. These provide us
with the in degree and out degree of the node.

For dendrograms (and a variety of other trees) we will want to be able
to find the edgeOrder and to change it.

The GXL description has edgeOrder. This seems like it could be useful
for drawing trees (and dendrograms) but it is only the outNodes that
need to be ordered.





